Metacognition and mathematics performance

### 2019-05-08 _analyze.Rmd
Revised manuscript

---
title: '`r paste("Experiment", params$experiment)`'
description: |
  Data analysis for Experiment `r params$experiment`.
author:
  - name: Matti Vuorre 
    affiliation: Columbia University
date: "`r Sys.Date()`"
output: 
  radix::radix_article:
    toc: true
    toc_depth: 2
    self_contained: false
params:
  experiment: 3
---

```{r setup, include=FALSE}
source("functions.R")
library(knitr)
library(glue)
library(ggbeeswarm)
library(summarytools)
library(tidyverse)
library(broom)
library(scales)
library(patchwork)
library(lme4)
library(afex)
library(emmeans)
opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  eval = TRUE,
  dev = c("png", "pdf"),
  fig.ext = c("png", "pdf"),
  fig.retina = 2
)
theme_set(theme_linedraw() + theme(panel.grid = element_blank()))
```

## Data description

@@

# Drop excluded subjects
dat <- filter(dat, !exclude)

@@

```{r metrics-summarise, results = 'asis'}
dat_metrics <- dat %>%
  nest(-subject) %>%
  mutate(res = map(data, metrics, ordinal = TRUE)) %>%
  unnest(res, .drop = TRUE)
# Summarize
dat_metrics %>% 
  select(-contains("s")) %>%
  descr(stats = c("mean", "sd", "min", "q1", "q3", "max", "n.valid", "pct.valid"), 
        round.digits = 2, style = "rmarkdown")
```

## Math performance

ANOVA of proportion correct on session. We use same exclusion rules as for the subject-level summaries, but at the session level. ANOVA only considers subjects who had observations at each session; thus we also fit a random intercept linear regression to include everyone.

```{r math-performance-sessions}
exclude_threshold <- 3

# Get gamma, confidence, and performance for everyone & sessions
dat_sessions <- dat %>%
  group_by(session, subject) %>%
  nest() %>%
  mutate(res = map(data, metrics)) %>%
  unnest(res)

# Create indicator for each of five exclusions rules for each subject. Variables are z-scored WITHOUT considering subjects with negative or NA gammas.
dat_sessions <- dat_sessions %>%
  mutate(
    e1 = is.na(gamma),
    e2 = gamma < 0,
    # Negative gammas to NAs so they don't impact next calculations
    gamma = ifelse(gamma < 0, NA, gamma),
    e3 = abs(scale(gamma)) > exclude_threshold,
    e4 = abs(scale(p)) > exclude_threshold,
    e5 = abs(scale(con_m)) > exclude_threshold
  ) %>%
  group_by(subject, session) %>%
  mutate(exclude = any(e1, e2, e3, e4, e5)) %>%
  ungroup() %>%
  filter(!exclude)
```

```{r math-performance-sessions-anova}
fit_aov_math_sessions <- aov_ez(
  id = "subject",
  dv = "p",
  data = dat_sessions,
  within = "session",
)
aov_n <- nrow(fit_aov_math_sessions$data$wide)
tidy(fit_aov_math_sessions$anova_table) %>% 
  kable(
    digits = 3, 
    caption = glue("G-G corrected ANOVA table; N = {aov_n})")
  )
```

```{r math-performance-sessions-anova2}
tidy(fit_aov_math_sessions$aov) %>%
  kable(
    digits = 3, 
    caption = glue("Uncorrected ANOVA table)")
  )
```

```{r math-performance-sessions-lme}
summary(lmer(p ~ session + (session || subject),
             data = dat_sessions))["coefficients"] %>%
  kable(digits = 3, caption = "lmer parameter summary")
```

```{r math-performance-sessions-fig, fig.cap = "Proportion correct across sessions."}
dat_sessions %>%
  ggplot(aes(session, p)) +
  # geom_point(shape = 1, position = position_jitter(h = 0, w = .1)) +
  geom_beeswarm(shape = 21) +
  stat_smooth(method = "lm") +
  stat_summary(fun.data = mean_cl_boot) +
  labs(
    x = "Session / test",
    y = "Proportion correct"
  )
```

@@

### Gamma by session

We used the same rejection rules for gammas at the session level.

```{r gamma-sessions-anova}
fit_aov_gamma_sessions <- aov_ez(
  id = "subject",
  dv = "gamma",
  data = dat_sessions,
  within = "session"
)
aov_n <- nrow(fit_aov_gamma_sessions$data$wide)
tidy(fit_aov_gamma_sessions$anova_table) %>% 
  kable(
    digits = 3, 
    caption = glue("G-G corrected gamma ANOVA table; N = {aov_n})")
  )
```

```{r gamma-sessions-anova2}
tidy(fit_aov_gamma_sessions$aov) %>%
  kable(
    digits = 3, 
    caption = glue("Uncorrected gamma ANOVA table)")
  )
```

```{r gamma-sessions-lme}
summary(lmer(gamma ~ session + (session || subject),
  data = dat_sessions))["coefficients"] %>%
  kable(digits = 3, caption = "Gamma lmer parameter summary")
```

```{r gamma-sessions-fig, fig.cap = "Gamma correlation across sessions."}
dat_sessions %>%
  ggplot(aes(session, gamma)) +
  geom_beeswarm(shape = 21) +
  stat_smooth(method = "lm") +
  stat_summary(fun.data = mean_cl_boot) +
  labs(
    x = "Session / test",
    y = "Gamma correlation"
  )
```

### Overall gamma

```{r gamma-overall}
dat_metrics %>%
  summarise(
    mean = mean(gamma, na.rm = T),
    N = n(),
    se = sd(gamma, na.rm = T) / sqrt(N)
  ) %>%
  kable(
    digits = 2,
    caption = "Summary of gamma correlations"
    )
```

```{r gamma-subject-ttest}
t.test(dat_metrics$gamma)
```

### Signal detection analysis

In calculating $d_a$, we added .5 to all cells. We calculate $d_a$ by obtaining intercept and slope of linear regression of z-scores in right panel above. We also obtained SDT metrics via ordinal regression to confirm.

@@

```{r resolution-scatterplots, fig.cap = "Average and 95pCIs gammas and das. Points = subjects."}
dat_metrics %>%
  select(subject, gamma, da) %>%
  gather(metric, value, gamma, da) %>%
  ggplot(aes(x = metric, y = value)) +
  scale_color_brewer(palette = "Set1") +
  geom_hline(yintercept = 0, lty = 2, size = .33) +
  geom_beeswarm(
    shape = 21,
    cex = 2,
    fill = "white"
  ) +
  stat_summary(
    fun.y = mean,
    geom = "point",
    size = 2.5
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = .1, size = .75
  ) +
  scale_y_continuous(breaks = pretty_breaks()) +
  facet_wrap("metric", scales = "free") +
  theme(
    axis.title.x = element_blank(),
    legend.position = "none"
  )
```

@@

```{r gamma-performance-plot, fig.cap = "Correlation between math performance and resolution."}
p_gamma_p <- dat_metrics %>%
  ggplot(aes(p, gamma)) +
  geom_vline(xintercept = .25, col = NA) + # Ensure limits
  geom_vline(xintercept = 1, col = NA) + # Ensure limits
  geom_hline(yintercept = 0, lty = 2, size = .33) +
  geom_point(shape = 21, fill = "white", alpha = .75) +
  geom_smooth(method = lm, col = "black", size = .75, alpha = .33) +
    scale_x_continuous(breaks = c(.25, .5, .75, 1), 
                     limits = c(.25, 1), expand = expand_scale(add = .01)) +
  labs(
    x = "Proportion correct",
    y = "Gamma"
  )
p_gamma_p
```

Pearson correlation between gamma and performance

```{r gamma-performance-test}
cor.test(dat_metrics$gamma, dat_metrics$p, method = "pearson")
```

Between $d_a$ and performance

```{r da-performance-test}
cor.test(dat_metrics$da, dat_metrics$p, method = "pearson")
```

## Calibration

### Overconfidence

```{r bias-calculate}
dat_bias <- dat %>%
  # Rescale confidence to 0-1 scale
  mutate(con_01 = ifelse(exp != 5, con6 / 5, con / 10)) %>%
  group_by(subject) %>%
  summarise(
    p = mean(accuracy, na.rm = T),
    con_01 = mean(con_01, na.rm = T)
  ) %>%
  mutate(bias = con_01 - p)
```

```{r bias-summary-test}
dat_bias %>%
  summarise(
    mean = mean(bias, na.rm = T),
    n = n(),
    se = sd(bias, na.rm = T) / sqrt(n),
    NAs = sum(is.na(bias))
  ) %>%
  kable(
    digits = 3,
    caption = "Summary of bias scores"
    )
t.test(dat_bias$bias)
```

```{r bias-plot, eval = F}
dat_bias %>%
  ggplot(aes(x = "bias", y = bias)) +
  scale_color_brewer(palette = "Set1") +
  geom_hline(yintercept = 0, lty = 2, size = .2) +
  geom_point(
    shape = 1,
    alpha = .5,
    size = .66,
    position = position_jitter(h = 0, w = .05)
  ) +
  stat_summary(
    fun.y = mean,
    geom = "point",
    size = 2.5
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = .33, size = .75
  ) +
  labs(y = "Bias") +
  theme(
    axis.title.x = element_blank(),
    legend.position = "none"
  )
```

### Bias - performance correlation

```{r bias-plot-performance, fig.cap = "Correlation between math performance and calibration bias."}
p_bias_p <- dat_bias %>%
  ggplot(aes(p, bias)) +
  geom_hline(yintercept = 0, lty = 2, size = .33) +
  geom_point(shape = 21, alpha = .75, fill = "white") +
  geom_smooth(method = lm, col = "black", size = .75, alpha = .33) +
  scale_x_continuous(breaks = c(.25, .5, .75, 1), 
                     limits = c(.25, 1), expand = expand_scale(add = .01)) +
  labs(
    x = "Proportion correct",
    y = "Mean confidence - proportion correct"
  )
p_bias_p
```

```{r bias-performance-correlate}
cor.test(dat_bias$bias, dat_bias$p)
```

## Publication figure

```{r figure-pub, fig.height = 3.4, fig.width = 6.6, fig.cap = "Figure for manuscript."}
(p_gamma_p + p_bias_p) *
  theme(aspect.ratio = 1) +
  plot_annotation(tag_levels = "A")
```

>>>>

---
title: '`r paste("Experiment", params$experiment)`'
author:
  - name: Matti Vuorre 
    affiliation: Columbia University
date: "`r Sys.Date()`"
output: 
  radix::radix_article:
    toc: true
    toc_depth: 2
    self_contained: false
params:
  experiment: 1
---

```{r setup, include=FALSE}
source("functions.R")
library(knitr)
library(ggbeeswarm)
library(tidyverse)
library(summarytools)
library(broom)
library(scales)
library(patchwork)
opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  eval = TRUE,
  dev = c("png", "pdf"),
  fig.ext = c("png", "pdf"),
  fig.retina = 2
)
theme_set(theme_linedraw() + theme(panel.grid = element_blank()))
```

## Descriptives

@@

@@

```{r metrics-summarise}
dat_metrics <- dat %>%
  nest(-subject) %>%
  mutate(res = map(data, metrics, ordinal = FALSE)) %>%
  unnest(res, .drop = TRUE)

descr(
  select(dat_metrics, n, p, gz, da, gamma), 
  stats = c("mean", "sd", "min", "max", "q1", "q3", "n.valid"),
  transpose = TRUE
) %>% 
  kable(digits = 2)
```

@@

### Gamma

```{r gamma-overall}
dat_metrics %>%
  summarise(
    N = n(),
    mean = mean(gz, na.rm = T),
    se = sd(gz, na.rm = T) / sqrt(N)
  ) %>% 
  kable(digits=3)
```

```{r gamma-subject-ttest}
t.test(dat_metrics$gz)
```

### da

@@

@@

```{r gamma-performance-plot, fig.cap = "Correlation between math performance and resolution."}
p_gamma_p <- dat_metrics %>%
  ggplot(aes(p, gz)) +
  geom_vline(xintercept = .25, col = NA) + # Ensure limits
  geom_vline(xintercept = 1, col = NA) + # Ensure limits
  geom_hline(yintercept = 0, lty = 2, size = .33) +
  geom_point(shape = 21, fill = "white", alpha = .75) +
  geom_smooth(method = lm, col = "black", size = .75, alpha = .33) +
  scale_x_continuous(
    breaks = c(.25, .5, .75, 1), 
    limits = c(.25, 1)
  ) +
  labs(
    x = "Proportion correct",
    y = "Gamma"
  )
```

Correlations between resolution and performance

```{r correlation-tests}
a <- dat_metrics %>% 
  select(subject, gz, da, p) %>% 
  gather(Metric, value, gz, da) %>% 
  group_by(Metric) %>% 
  nest() %>% 
  mutate(model = map(data, ~cor.test(.$value, .$p, method = "pearson"))) %>% 
  mutate(out = map(model, ~tidy(.))) %>% 
  unnest(out, .drop = TRUE) %>% 
  mutate(p.value = pvalue(p.value)) %>% 
  select(-alternative)

b <- dat_metrics %>% 
  select(subject, gamma, p) %>% 
  gather(Metric, value, gamma) %>% 
  group_by(Metric) %>% 
  nest() %>% 
  mutate(model = map(data, ~cor.test(.$value, .$p, method = "kendall"))) %>% 
  mutate(out = map(model, ~tidy(.))) %>% 
  unnest(out, .drop = TRUE) %>% 
  mutate(p.value = pvalue(p.value)) %>% 
  select(-alternative)

bind_rows(a, b) %>% 
  kable(digits = 2)
```

## Figure

```{r figure, fig.height = 3.4, fig.width = 6.6}
brinley(dat_metrics, p, gz) |
  brinley(dat_metrics, p, da) + 
  labs(y=expression(italic("d")[a]))
```

### 2019-05-08 index.Rmd
Revised manuscript

---
title: "Math Metacognition"
description: |
  Analyses of math metacognition data
date: "`r Sys.Date()`"
author:
  - name: Matti Vuorre 
    affiliation: Columbia University
output: 
  radix::radix_article:
    toc: true
    toc_depth: 1
    self_contained: false
site: radix::radix_website
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(summarytools)
opts_chunk$set(echo = FALSE)
```

# Confidence in mathematics data and analyses

Matti Vuorre <mv2521@columbia.edu>

Data and analyses discussed in Vuorre & Metcalfe (working title: "Confidence in mathematics"; in prep).

@@

# Reproduce

@@

# Data description

## Confidence scales

Experiments 1 - 3 used a Likert scale with the following labels:

>0 I completely guessed.  
1 I did more than guess, but I am not confident that my answer is correct.  
2 I have some idea how to do this problem, but I am not confident that my answer is correct.  
3 I am not confident that my answer is correct, nor am I confident that it is incorrect.  
4 I am fairly confident that my answer is correct.  
5 I am extremely confident that my answer is correct.

Experiment 5 used a 0-10 Likert scale. Experiments 4 and 6 were computer experiments that used a 0-1 VAS.

Because of these differences in scales, we have binned confidence ratings to six bins (3 experiments already had six-point ratings, so scaling is only done on experiments 4-6). The raw confidence ratings are variable `con`, binned ratings are variable `con6`.

## Variables

```{r, eval = T}
dat <- read_csv("data/merged.csv", guess_max = 25000)
head(dat) %>% kable()
```

`exclude` is a logical indicating whether the participant on that row met the exclusion criteria. `subject` labels are arbitrary numbers.

Number of subjects by experiment

```{r}
dat %>% 
  distinct(exp, subject, exclude) %>% 
  count(exp, exclude) %>%
  spread(exclude, n, sep = ": ") %>% 
  kable()
```

Number of items by experiment

```{r}
dat %>% 
  distinct(exp, item) %>% 
  count(exp) %>% 
  kable()
```

Number of items per subject (only non-excluded subjects are considered) by experiment

```{r}
dat %>% 
  filter(!exclude) %>% 
  distinct(exp, subject, item) %>% 
  count(exp, subject) %>% 
  group_by(exp) %>% 
  summarise(min = min(n),
            max = max(n),
            iqr_l = quantile(n, .25),
            iqr_u = quantile(n, .75)) %>% 
  kable()
```

>>>>

---
title: "Metacognition and mathematics performance"
date: "`r Sys.Date()`"
author:
  - name: Matti Vuorre 
    affiliation: Columbia University
output: 
  radix::radix_article:
    toc: true
    toc_depth: 1
    self_contained: false
  github_document:
    html_preview: false
site: radix::radix_website
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = FALSE)
```

Data and analyses related to Vuorre & Metcalfe (*"Metacognition about Mathematics is Excellent but the Relation between Performance and Resolution is Confounded by Guessing"*; in prep).

@@

## Reproduce

@@

## Data description

```{r, echo = TRUE, message = FALSE, collapse = TRUE}
dat <- read_csv("data/merged.csv", guess_max = 25000)
glimpse(dat, 1)
```

We binned confidence ratings to six bins (Experiments 1-3 already had six-point ratings, 4 & 6 were slider ratings, 5 had a 0-10 Likert scale). The raw confidence ratings are variable `con`, binned ratings are variable `con6`.
